Distinguishing Between Spin-Aligned and Isotropic Binary Black Hole Populations Using Gravitational Wave Observations

### 2017-03-27 paper/AlignedIsotropicGW.tex
All except for Cole's difficult-to-address comment about spin distribution discussion.

\affiliation{Department of Astronomy and Maryland Center for Theory
  and Computation, University of Maryland, College Park MD
  20742, United States}

@@

\begin{abstract}
  The first direct detections of \acp{GW} from merging \acp{BBH} open
  a unique window into the formation environment of massive stars and
  their compact remnants.  One promising signature of the formation
  environment is the angular distribution of the \ac{BH} spins;
  systems formed through dynamical interacions among already-compact
  objects are expected to have isotropic spin orientations while
  binaries formed from pairs of stars born together are more likely to
  have spins aligned with the binary orbit as a consequence of their
  joint evolution toward a \ac{BBH} system.  By considering existing
  \ac{GW} measurements of $\chieff$, the best-measured combination of
  spin parameters, in the three likely binary black hole detections
  GW150914, LVT151012, and GW151226, we show that the data already
  exhibit a $\OOneSigmaIsoAligned\sigma$ ($\OOneOddsIsoAligned$ odds
  ratio) preference for an isotropic angular distribution amongst a
  suite of models for the spin distribution.  By considering the
  effect of an additional 10 detections drawn from the various models
  in the suite we show that such an agumented data set would enable at
  least a $\OTwoSigmaIsoAlignedMin\sigma$ ($\OTwoOddsIsoAlignedMin$
  odds ratio) distinction between the isotropic and aligned models,
  and in most cases better than $5\sigma$ ($2.9 \times 10^{-7}$ odds
  ratio).  The existing data strongly suggest a dynamical formation
  secnario for the observed systems, and this will be confirmed (or
  overturned) confidently in the near future by subsequent
  observations.
\end{abstract}

@@

Following the detection of a merging \ac{BBH} system, \ac{PE} tools
\citep{2015PhRvD..91d2003V} compare model gravitational waveforms
\citep{2014PhRvD..89h4006P,2014PhRvD..89f1502T,2014PhRvL.113o1101H}
against the observed data to obtain a posterior distribution on the
parameter space that describes the compact binary source.  The spin
parameter with the largest effect on waveforms, and a correspondingly
tight constraint from the data, is a mass-weighted, aligned,
``effective spin,''
\begin{equation}
  \chieff = \frac{c}{GM} \left( \frac{\vec{S}_1}{m_1} + \frac{\vec{S}_2}{m_2}
  \right) \cdot \frac{\vec{L}}{\hat{L}} = \frac{1}{M} \left( m_1 a_1^z + m_2 a_2^z \right),
\end{equation}
where the $\hat{z}$ axis is aligned with the orbital angular momentum
of the binary, $m_{1,2}$ are the masses of the more-massive (1) and
less-massive (2) components, $M = m_1 + m_2$ is the total mass,
$\vec{S}_{1,2}$ are the spin angular momentum vectors of the black
holes in the binary, $\vec{L}$ is the orbital angular momentum vector,
and $a_{1,2}$ are the corresponding dimensionless spin parameters
\citep{2016PhRvL.116x1102A}.

@@

\begin{figure}
  \plotone{../plots/chi-eff-mock-posteriors}
  \caption{\label{fig:O1-posteriors} Approximate posteriors on
    $\chieff$ from the LIGO O1 observations in \citet{O1-BBH}.  We
    approximate the posteriors reported in \citet{O1-BBH} using
    Gaussians with the same median and 90\% credible interval as
    reported in \citet{O1-BBH}.  It is notable that none of the
    $\chieff$ posteriors extends to the high spins in
    electromagnetically-detected stellar-mass black hole systems
    \citep{2015PhR...548....1M}.}
\end{figure}

Small values of $\chieff$ can result from either intrinsically small
spins or larger spins whose direction is mis-aligned with the orbital
plane of the binary (i.e.\ spin vectors with with small
$z$-components).  Mis-alignment is capable of producing
\emph{negative} values of $\chieff$, however, while small, aligned
spins will always have $\chieff \geq 0$.  To quantify the degree of
support for these two alternate explanations of small $\chieff$ values
in the merging \ac{BBH} population, we compared the Bayesian evidence
for various simple models of the spin population using the \ac{GW}
data set.

Each of our models for the merging \ac{BBH} spin population assumes
that the merging black holes are of equal mass (this is marginally
consistent with the three observations \citep{O1-BBH}, and the
$\chieff$ distribution is not particularly sensitive to the mass ratio
between the merging objects---see Section \ref{sec:mass-ratio}).  We
assume that the population spin distribution factorises into a
distribution for the spin magnitude, $a$, and a distribution for the
spin angles.  Finally, we assume that the distribution of spins is
common to each component in a merging binary.  Choosing one of three
magnitude distributions
\begin{equation}
  \label{eq:magnitude-dists}
  p(a) = \begin{cases}
    2\left(1-a\right) & \textnormal{``low''} \\
    1 & \textnormal{``flat''} \\
    2\left(a-1\right) & \textnormal{``high''}
  \end{cases},
\end{equation}
and pairing with either an isotropic angular distribution or a
distribution that generates perfect alignment with the positive $z$
axis yields six different models for the $\chieff$ distribution.
These models are shown in Figure \ref{fig:chieff-distribution-models}.

@@

We fit hierarchical models of the three LIGO O1 observations using
these six different, zero-parameter population distributions (see
Section \ref{sec:hierarchical}).  We also fit three mixture models for
the population, where the spin magnitude distribution is fixed but the
angular distribution is a weighted sum of the isotropic and aligned
distribution.  The evidence, or marginal likelihood, for each of the
models is shown in Figure \ref{fig:O1-odds}.  For all three magnitude
distributions, the mixture models' posterior on the mixing fraction
peaked at 100\% isotropic, explaining why the zero-parameter,
pure-isotropic models are preferred over the single-parameter mixture
models for every magnitude distribution with this data set.  Not
surprisingly, given the small $\chieff$ values in the three detected
systems, the most-favoured model among those with an isotropic angular
distribution has the ``decreasing'' magnitude distribution; the most
favoured model among those with an aligned distribution also has the
``decreasing'' magnitude distribution.  The odds ratio between the
aligned and isotropic models is $\OOneOddsIsoAligned$, or
$\OOneSigmaIsoAligned\sigma$; thus the data strongly favour isotropic
spins among our suite of models.

@@

Estimates of the rate of \ac{BBH} coalescences give a reasonable
chance of 10 additional \ac{BBH} detections in the near future
\citep{O1-BBH,2016ApJ...833L...1A,2016ApJS..227...14A}.  Assuming an
additional 10 detections drawn from each of our six zero-parameter
models for the spin distribution, we find the odds ratios shown in
Figure \ref{fig:O2-predictions}.  We find that most scenarios with an
additional 10 detections allow to distinguish the correct angular
distribution with greater than $5\sigma$ ($2.9 \times 10^{-7}$ odds)
credibility; and in the worst case the distinction is
$\OTwoSigmaIsoAlignedMin\sigma$ ($\OTwoOddsIsoAlignedMin$ odds ratio).

@@

\section{Discussion}

Most of our resolving power for the spin angular distribution is a
result of the fact that our ``aligned'' models cannot produce
$\chieff < 0$ (see Figure \ref{fig:chieff-distribution-models}).  If
spins are intrinsically very small, with $a \lesssim 0.1$ always, then
it is no longer possible to resolve the negative spin with a small
number of observations.  As noted below, however, spins observed in
X-ray binaries are generally large, and most models of isolated binary
evolution do not produce systems with spins uniformly bounded to small
values.

@@

With this in mind, we performed our analysis using three straw-man
spin distributions for spin parameters from $a/M=0$ to 1: flat
($P(a/M)=1$); rising ($P(a/M)=2a/M$); and falling ($P(a/M)=2(1-a/M)$).
These are not meant to represent any particular physical model, and
indeed neither observations nor population synthesis codes can
authoritatively suggest {\it any} particular spin distribution.  This
does, however, allow us to see how sensitive the $\chieff$
distribution is to spin alignment models given uncertainties about the
spin magnitudes.

Although BH spin-orbit alignment is assumed in continuum flux
measurements \citep{2015PhR...548....1M}, some BH XRBs, including GRO
J1655-40 \citep{Martin:2008}, 4U 1543-47 \citep{MorningstarMiller:2014} and particularly V4641 Sgr \citep{Orosz:2001,Martin:2008b} appear to indicate that the micro
quasar jet, presumably aligned with the BH spin axis, is misaligned
with the orbit.  Moreover, initial stellar spins in massive binaries
have been observed to be misaligned \citep[e.g.,][]{Albrecht:2009}, though there are
opportunities for realignment during binary evolution, as described
below.

@@

The spin directions of binary black holes formed dynamically through
interactions in dense stellar environments \citep{SigurdssonHernquist:1993,PZMcMillan:2000,Rodriguez:2015,Stone:2016} 
are expected to be isotropic given the absence
of a preferred direction \citep[e.g.,][]{2016ApJ...832L...2R} and the
persistence of an isotropic distribution through post-Newtonian
evolution \citep{2004PhRvD..70l4020S}.

@@

The components of this formula are
\begin{itemize}
\item The GW (marginal) likelihood, $p\left(d \mid \chieff\right)$.
  ``Marginal'' because we are (implicitly) integrating over all
  parameters of the signal but $\chieff$.  Note that it is the
  likelihood, not the posterior that matters for the hierarchical
  analysis; if we are given posterior distributions or posterior
  samples, we need to re-weight to "remove" the prior and obtain the
  likelihood.
\item The population distribution for $\chieff$,
  $p\left( \chieff \mid \lambda \right)$.  This function can be
  parameterised by population-level parameters, $\lambda$.  (In the
  cases discussed above, there are no parameters for the population.)
\item The prior on the population-level parameters, $p(\lambda)$.
\end{itemize}
If we don't care about the individual event $\chieff$ paremetrs, we
can integrate them out, obtaining
\begin{equation}
  p\left( \lambda \mid \left\{ d^i \right\} \right) \propto \left[ \prod_{i=1}^{N_\mathrm{obs}} \int \dd \chieff^i \, p\left(d^i \mid \chieff^i \right) p\left( \chieff^i \mid \lambda \right) \right] p\left(\lambda\right).
\end{equation}
If we are given posterior samples of $\chieff^{ij}$ ($i$ labels the
event, $j$ labels the particular posterior sample) drawn from an
analysis using a prior $p\left( \chieff \right)$, then we can
approximate the integral by an re-weighted average of the population
distribution over the samples (here $p\left( \chieff^{ij} \right)$ is
the prior used to produce the posterior samples):
\begin{equation}
  p\left( \lambda \mid \left\{ d^i \right\} \right) \propto \left[ \prod_{i=1}^{N_\mathrm{obs}} \frac{1}{N_i} \sum_{j=1}^{N_i} \frac{p\left( \chieff^{ij} \mid \lambda \right)}{p\left( \chieff^{ij} \right)} \right] p\left(\lambda\right).
\end{equation}

>>>>

\affiliation{Department of Astronomy and Joint Space-Science
  Institute, University of Maryland, College Park, MD 20742--2421,
  United States}

@@

\begin{abstract}
  The first direct detections of \acp{GW} from merging \acp{BBH} open
  a unique window into the formation environment of massive stars and
  their compact remnants.  One promising signature of the formation
  environment is the angular distribution of the \ac{BH} spins;
  systems formed through dynamical interactions among already-compact
  objects are expected to have isotropic spin orientations whereas
  binaries formed from pairs of stars born together are more likely to
  have spins aligned with the binary orbit as a consequence of their
  joint evolution toward a \ac{BBH} system.  By considering existing
  \ac{GW} measurements of $\chieff$, the best-measured combination of
  spin parameters, in the three likely binary black hole detections
  GW150914, LVT151012, and GW151226, we show that the data already
  exhibit a $\OOneSigmaIsoAligned\sigma$ ($\OOneOddsIsoAligned$ odds
  ratio) preference for an isotropic angular distribution amongst a
  suite of models for the spin distribution.  By considering the
  effect of an additional 10 detections drawn from the various models
  in the suite we show that if all observations come from a single
  population such an agumented data set would enable at least a
  $\OTwoSigmaIsoAlignedMin\sigma$ ($\OTwoOddsIsoAlignedMin$ odds
  ratio) distinction between the isotropic and aligned models, and in
  most cases better than $5\sigma$ ($2.9 \times 10^{-7}$ odds ratio).
  The existing preference for a dynamical formation scenario for the
  observed systems will be confirmed (or overturned) confidently in
  the near future by subsequent observations.
\end{abstract}

@@

Following the detection of a merging \ac{BBH} system, \ac{PE} tools
\citep{2015PhRvD..91d2003V} compare model gravitational waveforms
\citep{2014PhRvD..89h4006P,2014PhRvD..89f1502T,2014PhRvL.113o1101H}
against the observed data to obtain a posterior distribution on the
parameter space that describes the compact binary source.  The spin
parameter with the largest effect on waveforms, and a correspondingly
tight constraint from the data, is a mass-weighted, aligned,
``effective spin,''
\begin{equation}
  \chieff = \frac{c}{GM} \left( \frac{\vec{S}_1}{m_1} + \frac{\vec{S}_2}{m_2}
  \right) \cdot \frac{\vec{L}}{\hat{L}} = \frac{1}{M} \left( m_1 a_1^z + m_2 a_2^z \right),
\end{equation}
where the $\hat{z}$ axis is aligned with the orbital angular momentum
of the binary, $m_{1,2}$ are the masses of the more-massive (1) and
less-massive (2) components, $M = m_1 + m_2$ is the total mass,
$\vec{S}_{1,2}$ are the spin angular momentum vectors of the black
holes in the binary, $\vec{L}$ is the orbital angular momentum vector,
and $0 \leq a_{1,2} \leq 1$ are the corresponding dimensionless spin
parameters \citep{2016PhRvL.116x1102A}.

@@

\begin{figure}
  \plotone{../plots/chi-eff-mock-posteriors}
  \caption{\label{fig:O1-posteriors} Approximate posteriors on
    $\chieff$ from the LIGO O1 observations in \citet{O1-BBH}.  We
    approximate the posteriors reported in \citet{O1-BBH} using
    Gaussians with the same median and 90\% credible interval as
    reported in \citet{O1-BBH}.  It is notable that none of the
    $\chieff$ posteriors extends to the high spin magnitudes inferred
    from several electromagnetically-detected stellar-mass black hole
    systems (see \citet{2015PhR...548....1M} for a summary of such
    measurements).}
\end{figure}

Small values of $\chieff$ can result from either intrinsically small
spins or larger spins whose direction is mis-aligned with the orbital
plane of the binary (i.e.\ spin vectors with small $z$-components).
Mis-alignment is capable of producing \emph{negative} values of
$\chieff$, however, while small, aligned spins will always have
$\chieff \geq 0$.  To quantify the degree of support for these two
alternate explanations of small $\chieff$ values in the merging
\ac{BBH} population, we compared the Bayesian evidence for various
simple models of the spin population using the \ac{GW} data set.

Each of our models for the merging \ac{BBH} spin population assumes
that the merging black holes are of equal mass (this is marginally
consistent with the three observations \citep{O1-BBH}, and the
$\chieff$ distribution is not particularly sensitive to the mass ratio
between the merging objects---see Section \ref{sec:mass-ratio}).  We
assume that the population spin distribution factorises into a
distribution for the spin magnitude, $a$, and a distribution for the
spin angles.  Finally, we assume that the distribution of spins is
common to each component in a merging binary.  Choosing one of three
magnitude distributions
\begin{equation}
  \label{eq:magnitude-dists}
  p(a) = \begin{cases}
    2\left(1-a\right) & \textnormal{``low''} \\
    1 & \textnormal{``flat''} \\
    2 a & \textnormal{``high''}
  \end{cases},
\end{equation}
and pairing with either an isotropic angular distribution or a
distribution that generates perfect alignment with the positive $z$
axis yields six different models for the $\chieff$ distribution.
These models are shown in Figure \ref{fig:chieff-distribution-models}.

@@

The distributions in Eq.\ \eqref{eq:magnitude-dists} are not meant to
represent any particular physical model, but rather to capture our
uncertainty about the spin magnitude distribution, discussed in detail
in Section \ref{sec:discussion}; neither observations nor population
synthesis codes can at this point authoritatively suggest {\it any}
particular spin distribution.  Our models, however, allow us to see
how sensitive the $\chieff$ distribution is to spin alignment given
uncertainties about the spin magnitudes.

We fit hierarchical models of the three LIGO O1 observations using
these six different, zero-parameter population distributions (see
Section \ref{sec:hierarchical}).  We also fit three mixture models for
the population, where the spin magnitude distribution is fixed but the
angular distribution is a weighted sum of the isotropic and aligned
distribution.  The evidence, or marginal likelihood, for each of the
models is shown in Figure \ref{fig:O1-odds}.  For all three magnitude
distributions, the mixture models' posterior on the mixing fraction
peaks at 100\% isotropic, which explains why the zero-parameter,
pure-isotropic models are preferred over the single-parameter mixture
models for every magnitude distribution with this data set.  Not
surprisingly, given the small $\chieff$ values in the three detected
systems, the most-favoured model among those with an isotropic angular
distribution has the ``decreasing'' magnitude distribution; the most
favoured model among those with an aligned distribution also has the
``decreasing'' magnitude distribution.  The odds ratio between the
aligned and isotropic models is $\OOneOddsIsoAligned$, or
$\OOneSigmaIsoAligned\sigma$; thus the data favour isotropic spins
among our suite of models.

@@

Estimates of the rate of \ac{BBH} coalescences give a reasonable
chance of 10 additional \ac{BBH} detections in the near future
\citep{O1-BBH,2016ApJ...833L...1A,2016ApJS..227...14A}.  Assuming an
additional 10 detections drawn from each of our six zero-parameter
models for the spin distribution, we find the odds ratios shown in
Figure \ref{fig:O2-predictions}.  We find that most scenarios with an
additional 10 detections allow the the correct angular distribution to
be distinguished with greater than $5\sigma$ ($2.9 \times 10^{-7}$
odds) credibility; and in the worst case the distinction is
$\OTwoSigmaIsoAlignedMin\sigma$ ($\OTwoOddsIsoAlignedMin$ odds ratio).

@@

\section{Discussion}
\label{sec:discussion}

Most of our resolving power for the spin angular distribution is a
result of the fact that our ``aligned'' models cannot produce
$\chieff < 0$ (see Figure \ref{fig:chieff-distribution-models}).  If
spins are intrinsically very small, with $a \lesssim 0.1$ always, then
it is no longer possible to resolve the negative spin with a small
number of observations.  As noted below, however, spins observed in
X-ray binaries are typically large, and most models of isolated binary
evolution do not produce systems with spins uniformly bounded to small
values.

@@

Although BH spin-orbit alignment is assumed in continuum flux
measurements \citep{2015PhR...548....1M}, some BH XRBs, including GRO
J1655-40 \citep{Martin:2008}, 4U 1543-47
\citep{MorningstarMiller:2014} and particularly V4641 Sgr
\citep{Orosz:2001,Martin:2008b} appear to indicate that the micro
quasar jet, presumably aligned with the BH spin axis, is misaligned
with the orbit.  Moreover, initial stellar spins in massive binaries
have been observed to be misaligned
\citep[e.g.,][]{Albrecht:2009,2014ApJ...785...83A}, though there are
opportunities for realignment during binary evolution, as described
below.

@@

The spin directions of binary black holes formed dynamically through
interactions in dense stellar environments
\citep{SigurdssonHernquist:1993,PZMcMillan:2000,Rodriguez:2015,Stone:2016}
are expected to be isotropic given the absence of a preferred
direction \citep[e.g.,][]{2016ApJ...832L...2R} and the persistence of
an isotropic distribution through post-Newtonian evolution
\citep{2004PhRvD..70l4020S,2007ApJ...661L.147B}.

@@

The components of this formula are
\begin{itemize}
\item The GW (marginal) likelihood, $p\left(d \mid \chieff\right)$.
  Here we use ``marginal'' because we are (implicitly) integrating
  over all parameters of the signal but $\chieff$.  Note that it is
  the likelihood rather than the posterior that matters for the
  hierarchical analysis; if we are given posterior distributions or
  posterior samples, we need to re-weight to "remove" the prior and
  obtain the likelihood.
\item The population distribution for $\chieff$,
  $p\left( \chieff \mid \lambda \right)$.  This function can be
  parameterised by population-level parameters, $\lambda$.  (In the
  cases discussed above, there are no parameters for the population.)
\item The prior on the population-level parameters, $p(\lambda)$.
\end{itemize}
If we do not care about the individual event $\chieff$ paremetrs, we
can integrate them out, obtaining
\begin{equation}
  p\left( \lambda \mid \left\{ d^i \right\} \right) \propto \left[ \prod_{i=1}^{N_\mathrm{obs}} \int \dd \chieff^i \, p\left(d^i \mid \chieff^i \right) p\left( \chieff^i \mid \lambda \right) \right] p\left(\lambda\right).
\end{equation}
If we are given posterior samples of $\chieff^{ij}$ ($i$ labels the
event, $j$ labels the particular posterior sample) drawn from an
analysis using a prior $p\left( \chieff \right)$, then we can
approximate the integral by a re-weighted average of the population
distribution over the samples (here $p\left( \chieff^{ij} \right)$ is
the prior used to produce the posterior samples):
\begin{equation}
  p\left( \lambda \mid \left\{ d^i \right\} \right) \propto \left[ \prod_{i=1}^{N_\mathrm{obs}} \frac{1}{N_i} \sum_{j=1}^{N_i} \frac{p\left( \chieff^{ij} \mid \lambda \right)}{p\left( \chieff^{ij} \right)} \right] p\left(\lambda\right).
\end{equation}
